---
title: "ML2 Projet : Calories prediction"
author: "Emma Tricquet/Louiza Mezine/Aqeel Hashmi/Yann L'Hotelier"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: show
---

## Overview

The primary goal is to develop a regression model that accurately estimates Calories (the target variable). By analyzing the relationship between exercise intensity, duration, and the user's physical profile, we can understand which factors
contribute most significantly to energy expenditure.

# Setup

```{r, MESSAGE = FALSE, WARNING = False}
# Loading necessary libraries
library(tidyverse)
library(randomForest)
library(xgboost)
library(caret)
library(corrplot)
library(Matrix)
library(Metrics)
library(ranger)   # Random Forest
library(gridExtra)
library(grid)
library(dplyr)    # Sorting / data manipulation
library(ggplot2)  # Visualization


# Fix random variables 
set.seed(42)

```


# Improvement of our initial data
## Data Loading 
```{r}
file_path <- "df_1000.csv"

# Read the dataset
df <- read.csv(file_path, header = TRUE, sep = ",", stringsAsFactors = FALSE)

df <- df_1000
dim(df)
```
```{r, message=true}
str(df)
```

## Train / Test Split (ONE TIME ONLY)
```{r }
train_index <- createDataPartition(df$Calories, p = 0.8, list = FALSE)

train_raw <- df[train_index, ]
test_raw  <- df[-train_index, ]
```


```{r Baseline Random Forest}
train_base <- train_raw %>%
  select(-User_ID) %>%
  mutate(Gender = ifelse(Gender == "male", 1, 0))

test_base <- test_raw %>%
  select(-User_ID) %>%
  mutate(Gender = ifelse(Gender == "male", 1, 0))

rf_base <- randomForest(
  Calories ~ .,
  data = train_base,
  ntree = 300
)

pred_base <- predict(rf_base, test_base)

MAE_base <- mean(abs(pred_base - test_base$Calories))
R2_base  <- cor(pred_base, test_base$Calories)^2

MAE_base
```


```{r}
R2_base
```

```{r Improved Feature Dataset}
train_improved <- train_raw %>%
  select(-User_ID) %>%
    mutate(
      Gender = ifelse(Gender == "male", 1, 0),
      BMI = Weight / (Height / 100)^2,
      Intensity = Heart_Rate * Duration,
      Temp_Above_Normal = Body_Temp - 37
    ) %>%
    select(-Weight, -Heart_Rate)

test_improved <- test_raw %>%
  select(-User_ID) %>%
    mutate(
      Gender = ifelse(Gender == "male", 1, 0),
      BMI = Weight / (Height / 100)^2,
      Intensity = Heart_Rate * Duration,
      Temp_Above_Normal = Body_Temp - 37
    ) %>%
    select(-Weight, -Heart_Rate)
```

```{r Improved Random Forest}
rf_improved <- randomForest(
  Calories ~ .,
  data = train_improved,
  ntree = 300,
  importance = TRUE
)

pred_improved <- predict(rf_improved, test_improved)

MAE_improved <- mean(abs(pred_improved - test_improved$Calories))
R2_improved  <- cor(pred_improved, test_improved$Calories)^2

MAE_improved
```

```{r}
R2_improved
```

```{r Baseline XGBoost}
train_xgb_base <- train_base
test_xgb_base  <- test_base

X_train_base <- train_xgb_base %>% select(-Calories)
y_train_base <- train_xgb_base$Calories

X_test_base  <- test_xgb_base %>% select(-Calories)
y_test_base  <- test_xgb_base$Calories

dtrain_base <- xgb.DMatrix(as.matrix(X_train_base), label = y_train_base)
dtest_base  <- xgb.DMatrix(as.matrix(X_test_base),  label = y_test_base)

params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  max_depth = 6,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_base <- xgb.train(
  params = params,
  data = dtrain_base,
  nrounds = 300,
  verbose = 0
)

pred_xgb_base <- predict(xgb_base, dtest_base)

MAE_xgb_base <- mean(abs(pred_xgb_base - y_test_base))
R2_xgb_base  <- cor(pred_xgb_base, y_test_base)^2

MAE_xgb_base
```

```{r}
R2_xgb_base
```

```{r Improved XGBoost}
X_train <- train_improved %>% select(-Calories)
y_train <- train_improved$Calories

X_test  <- test_improved %>% select(-Calories)
y_test  <- test_improved$Calories

dtrain <- xgb.DMatrix(as.matrix(X_train), label = y_train)
dtest  <- xgb.DMatrix(as.matrix(X_test),  label = y_test)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 300,
  verbose = 0
)

pred_xgb <- predict(xgb_model, dtest)

MAE_xgb <- mean(abs(pred_xgb - y_test))
R2_xgb  <- cor(pred_xgb, y_test)^2

MAE_xgb
```

```{r}
R2_xgb
```

```{r XGBoost Feature Importance}
importance_matrix <- xgb.importance(
  feature_names = colnames(X_train),
  model = xgb_model
)

xgb.plot.importance(importance_matrix)
```


## Final Comparison Table
```{r}
results <- data.frame(
  Model = c(
    "Baseline Random Forest",
    "Improved Random Forest",
    "Baseline XGBoost",
    "Improved XGBoost"
    ),
  
MAE = c(
  MAE_base,
  MAE_improved,
  MAE_xgb_base,
  MAE_xgb
  ),

R2 = c(
  R2_base,
  R2_improved,
  R2_xgb_base,
  R2_xgb
  )

)

results
```


# Applying Random Forest

## 2) Preprocessing

```{r preprocessing}
# Convert Gender to factor (categorical)
df$Gender <- factor(df$Gender)

# Remove the user identifier (not a useful predictor, risk of leakage)
df$User_ID <- NULL
```

## 3) Train / Validation / Test split

```{r split}
set.seed(42)

n <- nrow(df)

# Randomly shuffle indices to avoid biased splits if the dataset is ordered
idx <- sample(seq_len(n))

n_train <- floor(0.60 * n)  # 60% for training
n_val   <- floor(0.20 * n)  # 20% for validation
# The remainder (~20%) is the final test set

# Split the shuffled indices into 3 blocks
idx_train <- idx[1:n_train]
idx_val   <- idx[(n_train + 1):(n_train + n_val)]
idx_test  <- idx[(n_train + n_val + 1):n]

train_df <- df[idx_train, ]  # training set (fit models)
val_df   <- df[idx_val, ]    # validation set (hyperparameter selection)
test_df  <- df[idx_test, ]   # test set (final evaluation only)

dim(train_df)
dim(val_df)
dim(test_df)
```

## 4) Libraries, formula, and metrics

```{r libs-metrics}
target_formula <- Calories ~ .

# Metrics

rmse <- function(y_true, y_pred) {
  sqrt(mean((y_true - y_pred)^2))
}

mae <- function(y_true, y_pred) {
  mean(abs(y_true - y_pred))
}

# R-squared (coefficient of determination)
r2 <- function(y_true, y_pred) {
  1 - sum((y_true - y_pred)^2) / sum((y_true - mean(y_true))^2)
}
```

## 5) Baseline model (train → validation)

```{r baseline}
set.seed(42)

# p = number of predictors (all columns except target)
p <- ncol(train_df) - 1

# Course rule of thumb: mtry ~ sqrt(p)
# Here p = 7 => sqrt(7) ~ 2.64 => mtry = 2 (or 3)
baseline_model <- ranger(
  formula = target_formula,
  data = train_df,
  num.trees = 500,
  mtry = floor(sqrt(p)),
  min.node.size = 5
)

baseline_pred <- predict(baseline_model, data = val_df)$predictions

baseline_rmse <- rmse(val_df$Calories, baseline_pred)
baseline_mae  <- mae(val_df$Calories, baseline_pred)
baseline_r2   <- r2(val_df$Calories, baseline_pred)

cat("Baseline (Validation) — MAE:", baseline_mae,
    " RMSE:", baseline_rmse,
    " R2:", baseline_r2, "\n")
```

## 6) Hyperparameter tuning (grid search on validation)

We choose the configuration that **minimizes validation RMSE**, and we also report MAE and R².

```{r grid-search}
# Hyperparameter grid:
# - mtry: number of variables sampled as candidates at each split
# - min.node.size: controls tree depth / leaf size
# - num.trees: number of trees in the forest

grid <- expand.grid(
  mtry = 1:p,
  min.node.size = c(1, 3, 5, 10),
  num.trees = c(300, 500, 800)
)

nrow(grid)  # number of configurations

# Pre-allocate results (same number of rows as the grid)
results <- grid
results$RMSE_val <- NA_real_
results$MAE_val  <- NA_real_
results$R2_val   <- NA_real_

set.seed(42)
for (i in seq_len(nrow(grid))) {

  model_i <- ranger(
    formula = target_formula,
    data = train_df,
    num.trees = grid$num.trees[i],
    mtry = grid$mtry[i],
    min.node.size = grid$min.node.size[i]
  )

  pred_i <- predict(model_i, data = val_df)$predictions

  results$RMSE_val[i] <- rmse(val_df$Calories, pred_i)
  results$MAE_val[i]  <- mae(val_df$Calories, pred_i)
  results$R2_val[i]   <- r2(val_df$Calories, pred_i)
}

# Sort by RMSE (ascending)
results_sorted <- results %>% arrange(RMSE_val)

cat("\nTop 10 configurations (Validation, sorted by RMSE)\n")
print(head(results_sorted, 10))

# Best configuration
top1 <- results_sorted[1, ]
cat("\nBest configuration (Validation)\n")
print(top1)

# Save full tuning table
write.csv(results_sorted, "rf_tuning_results.csv", row.names = FALSE)
```

## 7) Final training (train + validation) and test evaluation

Important: to compare fairly on the **test set**, we refit:

- **baseline_final_model** with baseline hyperparameters on `(train + validation)`  
- **final_model** with tuned hyperparameters on `(train + validation)`  

Then evaluate both on `test_df`.

```{r final-train-test}
# Train on train + validation, then evaluate on test (fair comparison)
train_val_df <- bind_rows(train_df, val_df)

# --- Baseline refit on train+val, then evaluate on test ---
set.seed(42)
baseline_final_model <- ranger(
  formula = target_formula,
  data = train_val_df,
  num.trees = 500,
  mtry = floor(sqrt(p)),
  min.node.size = 5
)

baseline_test_pred <- predict(baseline_final_model, data = test_df)$predictions
baseline_rmse_test <- rmse(test_df$Calories, baseline_test_pred)
baseline_mae_test  <- mae(test_df$Calories, baseline_test_pred)
baseline_r2_test   <- r2(test_df$Calories, baseline_test_pred)

cat("\nBaseline (Test)\n")
cat("MAE_test :", baseline_mae_test,  "\n")
cat("RMSE_test:", baseline_rmse_test, "\n")
cat("R2_test  :", baseline_r2_test,   "\n")

# --- Tuned RF trained on train+val, then evaluate on test ---
set.seed(42)
final_model <- ranger(
  formula = target_formula,
  data = train_val_df,
  num.trees = top1$num.trees,
  mtry = top1$mtry,
  min.node.size = top1$min.node.size,
  importance = "impurity"   # needed to compute variable importance
)

final_test_pred <- predict(final_model, data = test_df)$predictions
final_rmse_test <- rmse(test_df$Calories, final_test_pred)
final_mae_test  <- mae(test_df$Calories, final_test_pred)
final_r2_test   <- r2(test_df$Calories, final_test_pred)

cat("\nTuned RF (Test)\n")
cat("MAE_test :", final_mae_test,  "\n")
cat("RMSE_test:", final_rmse_test, "\n")
cat("R2_test  :", final_r2_test,   "\n")

# Summary table
summary_compare <- data.frame(
  Model = c("Baseline RF (Test)", "Tuned RF (Test)"),
  MAE   = c(baseline_mae_test, final_mae_test),
  RMSE  = c(baseline_rmse_test, final_rmse_test),
  R2    = c(baseline_r2_test, final_r2_test)
)

cat("\nSummary (Baseline vs Tuned on Test)\n")
print(summary_compare)

write.csv(summary_compare, "rf_summary_baseline_vs_tuned.csv", row.names = FALSE)
```

## 8) Plot: Predicted vs True (test set)

This plot helps visually assess calibration and error patterns.  
A perfect model would lie on the diagonal line.

```{r plot-pred-vs-true}
plot_df <- data.frame(True = test_df$Calories, Pred = final_test_pred)

ggplot(plot_df, aes(x = True, y = Pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "Predicted vs True Calories (Test set) — Tuned Random Forest",
    x = "True Calories",
    y = "Predicted Calories"
  )
```

## 9) Variable importance (impurity-based)

This shows which features contribute most to decreasing impurity in splits.  
It is **useful for interpretation**, but remember: impurity importance can be biased toward variables with more potential split points.

```{r var-importance}
# Extract importance values
imp <- data.frame(
  feature = names(final_model$variable.importance),
  importance = as.numeric(final_model$variable.importance)
) %>% arrange(desc(importance))

cat("\nTop features (importance)\n")
print(head(imp, 10))

# Plot Top 10
ggplot(imp %>% slice_max(order_by = importance, n = 10),
       aes(x = reorder(feature, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Feature Importances — Tuned Random Forest",
    x = "Feature",
    y = "Importance (impurity)"
  )
```

## 10) Residual plot (test set)

Residuals = True − Pred.  
This plot helps detect systematic bias (e.g., under-predicting high values).

```{r residuals-plot}
res_df <- data.frame(
  True  = test_df$Calories,
  Pred  = final_test_pred,
  Resid = test_df$Calories - final_test_pred
)

ggplot(res_df, aes(x = True, y = Resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Residuals vs True Calories (Test set) — Tuned Random Forest",
    x = "True Calories",
    y = "Residual (True - Pred)"
  )
```

## 11) Save model and outputs

```{r save-outputs}
# Save the final tuned model
saveRDS(final_model, "final_rf_model.rds")

cat("\nFiles created:\n")
cat("- rf_tuning_results.csv (grid search results)\n")
cat("- rf_summary_baseline_vs_tuned.csv (baseline vs tuned comparison on test)\n")
cat("- final_rf_model.rds (final tuned random forest model)\n")
```

# Data presentation part
```{r}
file_path <- "calories.csv"

# Read the dataset
df <- read.csv(file_path, header = TRUE, sep = ",", stringsAsFactors = FALSE)
str(df)
df_1000 <- read.csv("df_1000.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)


par(mfrow = c(1,2))
hist(df$Calories, main = "Original", col = "grey")
hist(df_1000$Calories, main = "Sample - 1000", col = "grey")
```

# Applying XGBoost
```{r}
#install.packages(c("tidyverse", "xgboost", "Matrix", "caret", "Metrics"))
library(xgboost)
library(Matrix)
library(caret)
library(Metrics)
library(tidyverse)

#Data cleaning
df3 <- df_1000 %>%
  select(-User_ID) %>%              
  mutate(Gender = factor(Gender))

#Numerical matrix
X <- model.matrix(Calories ~ . - 1, data = df3)
y <- df3$Calories

#Split train / test
idx <- createDataPartition(y, p = 0.8, list = FALSE)

X_train <- X[idx, ]
y_train <- y[idx]

X_test  <- X[-idx, ]
y_test  <- y[-idx]

#DMatrix
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

#Parameters
params <- list(
  booster = "gbtree",
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,
  max_depth = 6,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 1
)

#Early stopping train
set.seed(42)

model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 5000,                         
  watchlist = list(train = dtrain, eval = dtest),
  early_stopping_rounds = 30,
  verbose = 1
)

#Prediction & evaluation
pred <- predict(model, dtest)

rmse_val <- rmse(y_test, pred)
mae_val  <- mae(y_test, pred)
r2_val   <- cor(y_test, pred)^2

rmse_val
mae_val
r2_val

#What are the most importants categories ?
imp <- xgb.importance(model = model)
print(imp)
xgb.plot.importance(imp)

#prediction
new <- data.frame(
  Gender = "male",
  Age = 22,
  Height = 178,
  Weight = 72,
  Duration = 30,
  Heart_Rate = 135,
  Body_Temp = 40.0
)

new$Gender <- factor(new$Gender, levels = levels(df3$Gender))

X_new <- model.matrix(~ . - 1, data = new)
pred_new <- predict(model, xgb.DMatrix(X_new))
pred_new

#length(y_test)
#length(pred)
```

```{r Graphs}

#Predicted vs True Calories - XGBoost
library(ggplot2)

df_pred <- data.frame(
  True = y_test,
  Predicted = pred
)

ggplot(df_pred, aes(x = True, y = Predicted)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  labs(
    title = "Predicted vs True Calories - XGBoost",
    x = "True Calories",
    y = "Predicted Calories"
  ) +
  theme_minimal()

#Residuals vs True Calories
df_res <- data.frame(
  True = y_test,
  Residual = y_test - pred
)

ggplot(df_res, aes(x = True, y = Residual)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black") +
  labs(
    title = "Residuals vs True Calories - XGBoost",
    x = "True Calories",
    y = "Residual (True - Predicted)"
  ) +
  theme_minimal()
```


# INTERPRETABLE MACHINE LEARNING

Understanding *why* the models make specific predictions is just as important as accuracy. 

Methods covered:
1. Permutation Feature Importance (PFI) -> model_parts()
2. Partial Dependence Plots (PDP)       -> model_profile(type = "partial")
3. Accumulated Local Effects (ALE)      -> model_profile(type = "accumulated")
4. SHAP values                          -> predict_parts(type = "shap")
# 1. SETUP AND LIBRARIES -------------------------------------------------------
```{r, message=FALSE, warning=FALSE}
library(DALEX)        # Core interpretability library
library(DALEXtra)     # Extension for wrappers
library(ggplot2)      # Visualization
library(dplyr)        # Data manipulation
library(gridExtra)    # Grid layouts
library(patchwork)    # For arranging plots
```

## 2. Create DALEX Explainers
To compare the models, we must wrap them in "Explainers". Note: We need a custom predict function for XGBoost because it expects a DMatrix, whereas DALEX passes data frames.

```{r}
# 1. Define the data to be used for explanation (Evaluation set)
# We use the test set (without the target column) for explanations
X_explain <- test_base %>% select(-Calories)
y_explain <- test_base$Calories
# 2. Custom Predict Function for XGBoost
predict_xgb_custom <- function(model, newdata) {
  # Convert dataframe to matrix for XGBoost
  newdata_mat <- xgb.DMatrix(as.matrix(newdata))
  predict(model, newdata_mat)
}

# 3. Create Explainers
# Assuming 'rf_improved' is your Random Forest model
# Assuming 'xgb_model' is your XGBoost model

explainer_rf <- DALEX::explain(
  model = rf_base,
  data = X_explain,
  y = y_explain,
  label = "Random Forest",
  verbose = FALSE
)

explainer_xgb <- DALEX::explain(
  model = xgb_base,
  data = X_explain,
  y = y_explain,
  predict_function = predict_xgb_custom,
  label = "XGBoost",
  verbose = FALSE
)

# Group explainers for easy iteration
explainers <- list(RF = explainer_rf, XGB = explainer_xgb)
```

## 3. MODEL Preformance (Residuals)
```{r}
# Calculate performance
mp_rf  <- model_performance(explainer_rf)
mp_xgb <- model_performance(explainer_xgb)

# Plot Reverse Cumulative Distribution of Residuals
plot(mp_rf, mp_xgb, geom = "boxplot") +
  ggtitle("Distribution of Residuals (Boxplot)") +
  theme_minimal()
```

# 4. Variable Importance (PFI)
```{r}
set.seed(42)

# Calculate PFI
pfi_rf  <- model_parts(explainer_rf, loss_function = loss_root_mean_square)
pfi_xgb <- model_parts(explainer_xgb, loss_function = loss_root_mean_square)

# Plot comparison
plot(pfi_rf, pfi_xgb, show_boxplots = FALSE) +
  ggtitle("Feature Importance (RMSE Loss)") +
  theme_minimal()
```

# 5. Feature Effects: Partial Dependence (PDP)
```{r}
# Select top 3 interesting features to plot
selected_features <- c("Duration", "Heart_Rate", "Age") 

# Make sure these columns exist in X_explain
features_to_plot <- intersect(selected_features, colnames(X_explain))

# Calculate PDP
pdp_rf  <- model_profile(explainer_rf, variables = features_to_plot, type = "partial")
pdp_xgb <- model_profile(explainer_xgb, variables = features_to_plot, type = "partial")

# Plot
plot(pdp_rf, pdp_xgb) +
  ggtitle("Partial Dependence Plots (PDP)") +
  theme_minimal()
```
This plot shows that Duration and Heart_Rate are the most influential predictors, both displaying a strong positive correlation with the average prediction. While both models follow similar trends, XGBoost is more sensitive to extreme values, whereas Random Forest is more conservative and tends to plateau (create plateaus) at high ranges.


# 6. Accumulated Local Effects (ALE)
```{r}

# Calculate ALE
ale_rf  <- model_profile(explainer_rf, variables = features_to_plot, type = "accumulated")
ale_xgb <- model_profile(explainer_xgb, variables = features_to_plot, type = "accumulated")

# Plot
plot(ale_rf, ale_xgb) +
  ggtitle("Accumulated Local Effects (ALE)") +
  theme_minimal()
```
This ALE plot confirms that Duration and Heart_Rate are the dominant features, showing a sharp increase in average prediction as their values rise. While both models agree on the general trends, XGBoost continues to predict higher values at the upper extremes of heart rate and duration, whereas Random Forest shows a clearer leveling off at high heart rates.

## 7.1 Local Interpretation: SHAP Values (with the biggest calories consumption)
```{r}
# 1. Select an interesting observation
# Let's pick the observation with the highest Duration in the test set
obs_index <- which.max(X_explain$Duration)
new_obs   <- X_explain[obs_index, , drop = FALSE]
real_val  <- y_explain[obs_index]

pred_rf  <- predict(rf_base, new_obs)
pred_xgb <- predict(xgb_base, as.matrix(new_obs))

# 2. Calculate SHAP values for Random Forest
shap_rf <- predict_parts(
  explainer = explainer_rf,
  new_observation = new_obs,
  type = "shap",
  B = 25  # Number of random paths
)

# 3. Calculate SHAP values for XGBoost
shap_xgb <- predict_parts(
  explainer = explainer_xgb,
  new_observation = new_obs,
  type = "shap",
  B = 25
)

# 4. Plot Comparison
p1 <- plot(shap_rf) +
  ggtitle(
    paste0(
      "SHAP: RF\nPredicted Cal: ", round(pred_rf, 1)
    )
  )

p2 <- plot(shap_xgb) +
  ggtitle(
    paste0(
      "SHAP: XGB\nPredicted Cal: ", round(pred_xgb, 1)
    )
  )

grid.arrange(
  p1, p2,
  ncol = 2,
  top = textGrob(
    paste("True Cal:", real_val),
    gp = gpar(fontsize = 16, fontface = "bold")
  )
)
```
Both models exhibit consistent directional trends for all features, showing a positive correlation across the entire observed range. However, there is a clear divergence of each feature contributions.

## 7.2 Local Interpretation: SHAP Values 
```{r}
# 1. Select an interesting observation
# Let's pick a random observation 
set.seed(42)
obs_index <- sample(seq_len(nrow(X_explain)), size = 1)
new_obs   <- X_explain[obs_index, , drop = FALSE]
real_val  <- y_explain[obs_index]

pred_rf  <- predict(rf_base, new_obs)
pred_xgb <- predict(xgb_base, as.matrix(new_obs))

# 2. Calculate SHAP values for Random Forest
shap_rf <- predict_parts(
  explainer = explainer_rf,
  new_observation = new_obs,
  type = "shap",
  B = 25  # Number of random paths
)

# 3. Calculate SHAP values for XGBoost
shap_xgb <- predict_parts(
  explainer = explainer_xgb,
  new_observation = new_obs,
  type = "shap",
  B = 25
)

# 4. Plot Comparison
p1 <- plot(shap_rf) +
  ggtitle(
    paste0(
      "SHAP: RF\nPredicted Cal: ", round(pred_rf, 1)
    )
  )

p2 <- plot(shap_xgb) +
  ggtitle(
    paste0(
      "SHAP: XGB\nPredicted Cal: ", round(pred_xgb, 1)
    )
  )


grid.arrange(
  p1, p2,
  ncol = 2,
  top = textGrob(
    paste("True Cal:", real_val),
    gp = gpar(fontsize = 16, fontface = "bold")
  )
)
```
We can find the same result here. 

## 7.2 Local Interpretation: SHAP Values 
```{r}
# Calculate Break Down for the best model (e.g., XGBoost)
bd_xgb <- predict_parts(
  explainer = explainer_xgb,
  new_observation = new_obs,
  type = "break_down"
)

plot(bd_xgb) + 
  ggtitle(paste("Break Down XGBoost | Real Value :", round(real_val, 1), "kcal")) +
  theme_minimal()
```


